\chapter{Questionnaires}\label{a:questionnaires}
Question types annotated with prefix and color: \colorbox{likert}{Likert-scale}, \colorbox{openq}{Open}, \colorbox{choice}{Choice}.

\section{Registration Questionnaire}
\begin{enumerate}
    \item \colorbox{choice}{C} How many years of programming experience do you have?
    \item \colorbox{choice}{C} How many years of professional programming experience do you have?
    \item \colorbox{likert}{L} How familiar are you with C and the C preprocessor?
    \item \colorbox{likert}{L} How familiar are you with Java?
    \item \colorbox{likert}{L} How familiar are you with merging?
    \item \colorbox{likert}{L} How familiar are you with diffing?
    \item \colorbox{likert}{L} How familiar are you with the UNIX patch tool?
    \item \colorbox{likert}{L} How familiar are you with pull requests?
    \item \colorbox{choice}{C} Which version control systems do you have experience with?
    \item \colorbox{choice}{C} Do you have previous experience with projectional editing?
\end{enumerate}

\section{Post-experiment Questionnaire}
\begin{enumerate}
    \item \colorbox{likert}{L} Integration with INCLINE is faster than in Eclipse.
    \item \colorbox{likert}{L} I make fewer mistakes in INCLINE than in Eclipse.
    \item \colorbox{likert}{L} Mistakes are easier to notice in INCLINE than in Eclipse.
    \item \colorbox{likert}{L} Mistakes are easier to fix in INCLINE than in Eclipse.
    \item \colorbox{likert}{L} INCLINE is a mature tool.
    \item \colorbox{likert}{L} Intention-based integration is not complex.
    \item \colorbox{likert}{L} The Keep intention is intuitive.
    \item \colorbox{likert}{L} The Remove intention is intuitive.
    \item \colorbox{likert}{L} The Implicit Keep/Remove intentions are intuitive.
    \item \colorbox{likert}{L} The Keep as Feature intention is intuitive.
    \item \colorbox{likert}{L} The Assign Feature intention is intuitive.
    \item \colorbox{openq}{O} Which intentions did you find the most useful?
    \item \colorbox{openq}{O} What are the advantages of using intentions for variant integration compared to manual integration with Eclipse?
    \item \colorbox{openq}{O} What are your perceived disadvantages of using intentions for variant integration compared to manual integration with Eclipse?
    \item \colorbox{openq}{O} Are there any possible improvements to intention-based integration or in particular INCLINE?
    \item \colorbox{openq}{O} How would you prefer to perform a variability-related integration?
\end{enumerate}