\chapter{Methodology}
Existing intentions into background. Cite authors title Under submission.

This chapter describes the intention language for variant integration, the prototype tool using the language, and the empirical evaluation of the language and tool.

\section{Intention-based variant integration}
Today, a developer performing a feature-related merge in a revision control system must make ad-hoc decisions on the final result, with respect to presence conditions over source code \cite{stanciulescu2016concepts}. This means that the concern of the developer is at the source-code level, dealing with fundamental (preprocessor) language constructs defining the presence conditions of particular code blocks. Previous studies show that variability is detrimental for program comprehension \cite{melo2016latin}, \cite{favre1997understanding}, \cite{ernst2002preprocessor}, \cite{abal2014variability}, \cite{medeiros2015preprocessor} \todo{Possibly also cite Ribeiro 2014}. We note that feature integration is a concern on the architectural level, while the merge is performed line-by-line on the source-code level. The mismatch between the architectural concern and the abstraction level on which it is carried out ensures that the integration (or merge) process is time consuming and has a high risk of introducing defects.

We propose to correct the mismatch of abstraction levels by introducing a domain specific language capturing the semantics of the feature integration. In particular, the language describes the outcome that the developer wants to achieve. Examples could be ``keep this functionality'' or ``remove this functionality''. We subsequently refer to these concepts simply as the \textit{intentions} of the developer. Each intention carries a mapping to actual operations that are performed on the source code. The semantics of the intentions language ensures the integrity of the underlying code upon which operations are applied to. This intentions language is the basis for creating a tool to aid the integration process structurally \cite{mens2002}, \cite{apel2011}. With proper tool support, the integration process can become more efficient from time consumption, risk, and quality perspectives. Indeed, tool support has shown to reduce variability errors in previous studies \cite{ribeiro2014emergent}.

The intentions and their underlying transformations can be described using choice calculus, a formal notation for variational software, independent of programming language \cite{erwig2011choice}, \cite{walkingshaw2012choice}. In the implementation, this can be realized as conditional compilation (C preprocessor) or design patterns (object-oriented languages). Intentions could also be mapped to the source-code operations identified by Stanciulescu et al in \cite{stanciulescu2016concepts}

Gousios et al. point out that an important improvement for the merging process is tool support in terms of work prioritization and estimated time for merging \cite{gousios2015}, and a prerequisite for the capability of migrating \textit{clone-and-own} to an integrated platform is having proper merge support \cite{antkiewicz2014flexible}.

We now present the intentions and workflow offered by out prototype tool INCLINE/bacon. The work has been carried out jointly between Max Lillack, Ștefan St\u{a}nciulescu, Wilhelm Hedman, Thorsten Berger, and Andrzej W\k{a}sowski.

\subsection{INCLINE/bacon Intentions}
\begin{itemize}
    \item{Keep}
    \item{Remove}
    \item{KeepAsFeature}
    \item{Exclusive}
    \item{AssignFeature}
    \item{Order}
    \item{Postpone}
    \item{++ implicit Keep and Remove}
\end{itemize}

\subsection{Workflow}
\begin{enumerate}
    \item Input two variant files, $(v, v')$. 
    \item Generate XML-based AST of each.
    \item Create initial/intermediary integrated model by merging ASTs of $(v, v')$.
    \item Simplify boolean conditions.
    \item Open in BACONTOOL.
    \item Apply intentions until satisfied.
    \item Serialize AST to text.
\end{enumerate}


\section{Data collection}
Two methods for collecting data from open-source ecosystems have been used, one for gathering a ground truth for how variant integration is achieved through merging, and one for procuring variants for hypothetical but realistic \todo{realistic or authentic?} integration scenarios. We denote the output of the former method \textit{merge examples} and the output of the latter method \textit{integration scenarios}. The merge examples are used for understanding the context and execution of the variant-integrating merge. They can also be replayed for validation, since they represent the ground truth outcome of the two merge parents. As such, the merge examples serve as an important validation tool. However, since the merge examples are extracted from a three-way merging tool, they are suboptimal for replaying in BACONTOOL, since three-way information is  not available in that environment. To this end, we use the integration scenarios, which are free from three-way contamination, but instead have no ground truth.

\subsection{Collecting ground truth merge examples}
To identify and validate the integration intentions we analyze variant-related merges and extract common patterns. We begin by retrieving all merges, and discard those that merged cleanly, thereby retaining those that merged with conflicts. Merged with conflicts signifies a syntactic conflict which is an indicator of potential feature integration. In order to investigate further, for each of these merges, we save the entire source tree as it was a) after the merge, i.e. the \textit{result}, b) in the first parent, i.e. the \textit{base}\todo{Change terminology to avoid confusion with git merge-base}, and c) in the second parent, the \textit{remote}. We call this triple of source trees the \textit{related artifacts} of the merge, the relationship between its parts is shown in Figure \ref{intentions:mergespace}. Retaining and inspecting the source trees of the parents allows for insights to be gathered, and in particular the ability to replay the merge in order to reach the known result. Note that the search space, the entire set of merges encompasses \textit{all} merges, that is, merges that occurred in a fork but were subsequently merged into the mainline are also included. We do \textit{not} use the \texttt{-{}-first-parent} option to Git, which would discard such further merges residing in recursive parents, and instead use only the first parent when finding merges. The merges that are included by this broader search scope originate either from ordinary feature merging or from pull requests where the branch of the requester is out of date.%\todo{Figure showing ABL feature integration and Deltabot feature integration.}

When the related artifacts from each merge have been collected, we inspect them and discard merges with any of the following properties: 
a) all conflicts occurred in non-source-code artifacts, b) some artifacts cannot be compiled, c) there are only whitespace changes between the merge parents, or d) contain some project-specific uninteresting changes (see Execution below for examples).

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
        \tikzset{every node/.append style={font=\sffamily}}
        \tikzstyle{linez} = [thick]
        \tikzstyle{commit} = [shape=circle,draw=black, minimum size=1cm, linez]
        \tikzstyle{ancestor} = [draw=gray, dashed]
        \tikzstyle{messager} = [xshift=1cm, yshift=-0.6cm, anchor=west]
        \tikzstyle{messagel} = [xshift=-1cm, yshift=-0.6cm, anchor=east]
    
        \node[commit, double, label={[messager]Result}] (res) at (0,0) {$m$};
        \node[commit, label={[messager]Base}] (ba) at (0,2) {$p_1$};
        \node[commit, label={[messagel]Remote}] (re) at (-2,2) {$p_2$};
        \node[commit, ancestor, label={[messager, color=gray]Common ancestor}] (an) at (0,4) {};
        
        \draw[thick, decoration={brace,mirror,raise=5pt},decorate] (3,-0.5) -- node[right=6pt] {Related artifacts} (3, 2.5);
        
        \draw[thick,->] (-6,4.5) to node[pos=0.5, xshift=0.2cm, rotate=-90] {$time$} (-6,-0.5);
        
        \draw [->, -latex, linez] (res) -- (ba);
        \draw [->, -latex, linez] (res) -- (re);
        \draw [->, -latex, ancestor, linez] (re) -- (an);
        \draw [->, -latex, ancestor, linez] (ba) -- (an);
        % NOTE: There is a bug in tikz with -latex arrows and colors. The arrows should therefore be drawn last, otherwise the colors of the arrowheads can mess up.
        
    \end{tikzpicture}
    \caption{Merge commit context. The merge commit $m$ has two parents, $p_1$ and $p_2$. The entire state of the source tree is gathered, as they appear in these commits, denoted \textit{result}, \textit{base}, and \textit{remote}. This triple of source tree sets constitute the \textit{related artifacts} of the commit $m$.}
    \label{intentions:mergespace}
\end{figure}

\subsection{Creating authentic merging tasks}
We select a number of forks that have some \textit{interesting} changes \cite{stanciulescu2015}. It does not matter whether they are still actively maintained or not, because we set the scenario at the \head of the fork, and then compare with either a) the latest common ancestor commit \texttt{\#ancestor} in fork and mainline, or b) an arbitrary number of commits after \texttt{\#ancestor} in the mainline. We create scenarios by partially replaying conflict merges, based on their variability-related diffs.

Done on BusyBox + TBD: Vim.

\section{Internal evaluation}
\subsection{Method}
We both replay merge examples and merge authentic scenarios. Each participant receives a task, and performs it in both Eclipse CDT and BACONTOOL. The number of actions for each task in each editor is saved for analysis, as is the resulting output file. Each tasks consists of a number of input files from the two variants being integrated, together with a textual description of what the integration goal is.

\subsection{Execution}
Three developers perform 3-4 tasks, at least one of each kind.

\subsection{Lessons learned}
\begin{enumerate}
    \item Provide expected output source file. Interpreting instructions is not what the experiment is about, it's about editing efficiency.
    \item Shorten files (6000 lines is too much, Vim even has 26k). 100-200 should be enough.
    \item Subjects will require training in BACONTOOL.
\end{enumerate}

\section{Controlled experiment}
This sections desrcibes the experimental design and setup.
%Designed based on Melo et al \cite{melo2016latin} (also the subsequent paper from 2017). % -- it has really good related work that can be used to show that the preprocessor (and configurations) are often erroneous.

\subsection{Objective}
To establish the beneficence of of the intention-based approach in general, and the intention-based integration tool in particular, we evaluate the tool using subject programs from other sources than Marlin, since Marlin was used to elicit the intentions. The purpose of this controlled experiment is to answer the following research questions:
\begin{enumerate}[label={Q\arabic*}]
    \setcounter{enumi}{1}
    \item \RQB
    \item \RQC
\end{enumerate}

\subsection{Treatments}
In order to study the beneficence of our intention-based approach, we let each participant solve one task using an ordinary unstructured merge tool, and one task using BACONTOOL. The two treatment levels are thus \textit{unstructured merging tool} \texttt{UNS} and \textit{Incline-bacon} \texttt{MPS}.

\subsection{Subjects}
We prepare two programs to be used in the experiment, using the process for collecting authentic variant integration scenarios, \po, based on \busybox, and \pt, based on \vim. These two programs are condensed versions of the entire changeset of a particular fork variant, for brevity and comprehension during the experiment. Table \ref{method:charac} lists the characteristics of the subject programs. \textit{Chunks and excluded chunks} shows the number of selected chunks from the full changeset to be in the program, and the excluded count. \textit{Average Included (Excluded) Chunk Length} (AICL/AECL) is the average length of each included chunk and excluded chunk. This should be understood as selecting the most suited changes across the files, and placing them in a single file, in order to not waste valuable time for the developers on long blocks without any changes, as to keep the programs brief.

(LOC might not be applicable, since we have two input files)

We choose \busybox~ and \vim~ as representative subjects because they are highly-configurable systems, and have been subjects in previous studies \cite{berger2013study}, \cite{liebig2010preprocessor}, \cite{liebig2011discipline}.

\begin{table}[h]
    \centering
    \caption{Characteristics of the tasks \po~and \pt.}
    \label{method:charac}
    \begin{tabular}{c l c c c c c}
    \hline
    \hline
        \textbf{Prg} & \textbf{Origin} & \textbf{LOC} & \textbf{\#chunks} & \textbf{AICL} & \textbf{\#excl. chunks} & \textbf{AECL} \\\hline
        \po & \busybox & - & - & - & - & -\\\hline
        \pt & \vim & - & - & - & - & - \\
        \hline
        \hline
    \end{tabular}
\end{table}

\subsection{Design}
The experiment is a 2$\times$2 \textit{within-subject/within-group} \todo{Which is it?} Latin square design \cite{box}. Each developer performs two tasks with the treatment order and program order randomized in order to reduce learning effects. Table \ref{method:ls1} shows the base Latin square.
Since developers are assigned at random, no further permutation of the treatments in the 2$\times$2 Latin square is required, since the two other possible combinations would be redundant with respect to program order and treatment.

\begin{table}[h]
\centering
\caption{Latin Square (2$\times$2).}
\label{method:ls1}
\begin{tikzpicture}
    \node[minimum size=1cm,draw=black] (sba) at (0,0) {\texttt{MPS}};
    \node[minimum size=1cm,draw=black] (sbb) at (1,0) {\texttt{UNS}};
    \node[minimum size=1cm,draw=black] (saa) at (0,1) {\texttt{UNS}};
    \node[minimum size=1cm,draw=black] (sab) at (1,1) {\texttt{MPS}};
    \node[minimum size=1cm] (ld1) at (-1,1) {\texttt{D1}};
    \node[minimum size=1cm] (ld2) at (-1,0) {\texttt{D2}};
    \node[minimum size=1cm] (lp1) at (0,2) {\texttt{P1}};
    \node[minimum size=1cm] (lp2) at (1,2) {\texttt{P2}};
    
    \node[minimum size=2.15cm,draw=black] (square) at (0.5,0.5) {};
    \node[] (lprog) at (0.5, 2.5) {\textit{programs}};
    \node[rotate=90] (ldevs) at (-1.75, 0.5) {\textit{developers}};
\end{tikzpicture}
\end{table}

\subsection{Execution}
TBD.
